```{r}
library(dplyr)
library(scatterplot3d)
library(plotly)
library(cars)
library(caret)
library(tidyr)
set.seed(123)
```

Lets convert the polar coordinates to cartesian
```{r}
data <- read.csv("./data/star_data_coords.csv")
df <- na.omit(data)

colnames(df) <- c("d", "tmp", "mh", "logg", "mag", "ra", "dec", "par")

distance <- 1/(df$par/1000) #divide the parallax by 1000 to convert milliarcseconds to arcseconds

# Convert RA and Dec to radians
ra_rad <- df$ra*pi/180
dec_rad <- df$dec*pi/180

# Calculate x, y, z coordinates
x <- distance * cos(dec_rad) * cos(ra_rad)
y <- distance * cos(dec_rad) * sin(ra_rad)
z <- distance * sin(dec_rad)

coords <- data.frame(x = x, y = y, z = z)

par(bg = "grey10",
    mar = c(5, 5, 2, 2))
scatterplot3d(x, y, z, pch = 16, color = "yellow", 
              xlim = c(-1e6, 1e6), ylim = c(-5e6, 5e6), zlim = c(-4e6, 4e6), 
              cex.symbols = 0.3,
              col.main = "white",   # Title text color
              col.lab = "white",    # Axis labels color
              col.axis = "white",
              main = "Star distance from Earth (pc)")

p <- plot_ly(x = x, y = y, z = z, type = "scatter3d", mode = "markers", marker = list(size = 1.5, color = "yellow"))

# Modify axis limits for 3D plot
p <- p %>% layout(
  scene = list(
    bgcolor = 'black',
    xaxis = list(range = c(-1e6, 1e6)),  # Set x-axis limits
    yaxis = list(range = c(-5e6, 5e6)),  # Set y-axis limits
    zaxis = list(range = c(-4e6, 4e6))   # Set z-axis limits
  )
)
p
```
  

Take errors into account
```{r}
data <- read.csv("./data/star_data_e.csv")
df <- na.omit(data)

colnames(df) <- c("tmp", "mh", "logg", "mag", "ra", "dec", "par", "ra_e", "dec_e", "par_e")



ggplot(df, aes(x = ra_e)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of RA Errors", x = "Error", y = "Frequency") +
  theme_minimal()

ggplot(df, aes(x = dec_e)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Dec Errors", x = "Error", y = "Frequency") +
  theme_minimal()

ggplot(df, aes(x = par_e)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Parallax Errors", x = "Error", y = "Frequency") +
  theme_minimal()



df_no_error <- df |>
  filter(ra_e <0.075, dec_e <0.075, par_e < 0.075)

distance <- 1/(df$par/1000) #divide the parallax by 1000 to convert milliarcseconds to arcseconds

# Convert RA and Dec to radians
ra_rad <- df_no_error$ra*pi/180
dec_rad <- df_no_error$dec*pi/180

# Calculate xyz coordinates
x <- distance * cos(dec_rad) * cos(ra_rad)
y <- distance * cos(dec_rad) * sin(ra_rad)
z <- distance * sin(dec_rad)

coords <- data.frame(x = x, y = y, z = z)

par(bg = "grey10",
    mar = c(5, 5, 2, 2))

scatterplot3d(x, y, z, pch = 16, color = "yellow", 
              xlim = c(-1e6, 1e6), ylim = c(-5e6, 5e6), zlim = c(-4e6, 4e6), 
              cex.symbols = 0.3,
              col.main = "white",   # Title text color
              col.lab = "white",    # Axis labels color
              col.axis = "white",
              main = "Star distance from Earth (pc)")

p <- plot_ly(x = x, y = y, z = z, type = "scatter3d", mode = "markers", marker = list(size = 1.5, color = "yellow"))

# Modify axis limits for 3D plot
p <- p %>% layout(
  scene = list(
    bgcolor = 'black',
    xaxis = list(range = c(-1e6, 1e6)),  # Set x-axis limits
    yaxis = list(range = c(-5e6, 5e6)),  # Set y-axis limits
    zaxis = list(range = c(-4e6, 4e6))   # Set z-axis limits
  )
)
p
```

  
  
  
```{r}
data <- read.csv("./data/star_data_coords.csv")
df <- na.omit(data)

colnames(df) <- c("d", "tmp", "mh", "logg", "mag", "ra", "dec", "par")
df$d <- 1000/df$par

linear <- lm(mh ~ par + tmp + logg + mag, data = df)
summary(linear)
```
***Talk about boring things***

Wow so many significant variables, lets see why
```{r}
#check variance inflation factor (VIF)
vif <- vif(linear)
vif

#normalize data
linear_data <- df |>
  select(mh, par, tmp, logg, mag)

preproc_l <- preProcess(linear_data, method = c("center", "scale"))
linear_data_norm <- predict(preproc_l, linear_data)

linear_norm <- lm(mh ~ par + tmp + logg + mag, data = linear_data_norm)
summary(linear_norm)

#lower sample size
subset_linear <- linear_data_norm |> sample_n(500)
linear_small <- lm(mh ~ par + tmp + logg + mag, data = subset_linear)
summary(linear_small)
```

***Talk about how to pick the right sample size***
  
Try to find how many datapoints are within the milkyway
```{r}
ggplot(df, aes(x = d)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(title = "Density Plot of Distances", x = "Distance (pc)", y = "Density") +
  xlim(0, 100000)

ggplot(df, aes(x = d)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(title = "Density Plot of Distances", x = "Distance (pc)", y = "Density") +
  xlim(0, 5.5e6)
```

Let's remove all data outside of 25,000 parsecs to get only stars within the milkyway
```{r}
data <- read.csv("./data/star_data_e.csv")
df <- na.omit(data)

colnames(df) <- c("tmp", "mh", "logg", "mag", "ra", "dec", "par", "ra_e", "dec_e", "par_e")
df$d <- 1000/df$par

df <- df |> filter(abs(d) < 25000)

df_no_error <- df |>
  filter(ra_e <0.05, dec_e <0.05, par_e < 0.05)

distance <- 1/(df_no_error$par/1000) #divide the parallax by 1000 to convert milliarcseconds to arcseconds

# Convert RA and Dec to radians
ra_rad <- df_no_error$ra*pi/180
dec_rad <- df_no_error$dec*pi/180

# Calculate xyz coordinates
x <- distance * cos(dec_rad) * cos(ra_rad)
y <- distance * cos(dec_rad) * sin(ra_rad)
z <- distance * sin(dec_rad)

b <- plot_ly(x = x, y = y, z = z, type = "scatter3d", mode = "markers", marker = list(size = 1.5, color = "yellow"))

# Modify axis limits for 3D plot
b <- b %>% layout(
  scene = list(
    bgcolor = 'black',
    xaxis = list(range = c(-2e4, 0), title = "x (pc)"),  # Set x-axis limits
    yaxis = list(range = c(-2.5e4, 1e4), title = "y (pc)"),  # Set y-axis limits
    zaxis = list(range = c(-2e4, 1e3), title = "z (pc)")   # Set z-axis limits
  )
)
b

#Make a pretty graph with earth in it
coords <- data.frame(x, y, z,
  source = rep("dataset", length(x)),
  size = rep(1.5, length(x))# Assign the label 'dataset' to each point
)

df <- df |> filter(abs(d) > 1000)
# Define the extra point (for example, at coordinates 6, 2, 3)
earth <- data.frame(
  x = 0, 
  y = 0, 
  z = 0,
  source = "earth",
  size = 4
  # Label the extra point as 'extra_point'
)

df_with_earth <- rbind(coords, earth)

p <- plot_ly(data = df_with_earth, x = ~x, y = ~y, z = ~z, type = "scatter3d", mode = "markers", marker = list(size = ~size), color = ~source, colors = c("yellow", "deepskyblue2"))

p <- p %>% layout(
  scene = list(
    bgcolor = 'black',
    xaxis = list(range = c(-2e4, 2e4)),  # Set x-axis limits
    yaxis = list(range = c(-2.5e4, 2.5e4)),  # Set y-axis limits
    zaxis = list(range = c(-2e4, 2e4))   # Set z-axis limits
  )
)
p
```

TEST GMT ERROR INTERVALS
```{r}
data <- read.csv("./data/star_data.csv")
df <- na.omit(data)

colnames(df) <- c("tmp", "mh", "logg", "mag", "ra", "dec", "par", "ra_e", "dec_e", "par_e", "logg_low", "logg_up", "mh_low", "mh_up","tmp_low", "tmp_up")

#make confidence intervals for mh and logg
df$int_mh <- df$mh_up - df$mh_low
df$int_logg <- df$logg_up - df$logg_low
df$int_tmp <- df$tmp_up - df$tmp_low

#calculate distance
df$d <- 1/(df$par/1000)

#remove upper and lower bound columns for space efficiency
df <- df |> 
  select(-mh_low, -mh_up, -logg_up, -logg_low, -tmp_up, -tmp_low) |>
  filter(d<25000)

df_errors <- df |> select(int_mh, int_logg, int_tmp)

df_errors <- df_errors %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(df_errors, aes(x = value, color = variable, fill = variable)) +
  geom_density(alpha = 0.3) +
  labs(x = "Error Interval", y = "Density", title = "Density of gmt error intervals") +
  xlim(0, 5) +
  theme_minimal()

summary(df$int_mh)
summary(df$int_tmp)
summary(df$int_logg)
```

BUILD CODE FOR GOOD DATASET
```{r}
data <- read.csv("./data/star_data.csv")
df <- na.omit(data)

colnames(df) <- c("tmp", "mh", "logg", "mag", "ra", "dec", "par", "ra_e", "dec_e", "par_e", "logg_low", "logg_up", "mh_low", "mh_up","tmp_low", "tmp_up")

#make confidence intervals for mh and logg
df$int_mh <- df$mh_up - df$mh_low
df$int_logg <- df$logg_up - df$logg_low
df$int_tmp <- df$tmp_up - df$tmp_low

#calculate distance
df$d <- abs(1000/df$par)

#convert apparent mag to absolute
df$mag <- df$mag - 5*log10(df$d) + 5

#calculate luminocity
l_sun <- 3.828e26
df$l <- l_sun * 10^((4.83-df$mag)/2.5)

#extract coordinate data and corresponding errors
coords <- df |>
  filter(abs(d) < 25000, int_tmp < 24, int_mh < 0.0476, int_logg < 0.025) |>
  select(ra, dec, par, par_e, dec_e, ra_e, d)

coords$ra_rad <- coords$ra*pi/180
coords$dec_rad <- coords$dec*pi/180

coords$x <- coords$d * cos(coords$dec_rad) * cos(coords$ra_rad)
coords$y <- coords$d * cos(coords$dec_rad) * sin(coords$ra_rad)
coords$z <- coords$d * sin(coords$dec_rad)

coords <- coords |> filter(z<0)
#remove used columns for space efficiency
df <- df |> 
  filter(abs(d) < 25000, int_tmp < 24, int_mh < 0.0476, int_logg < 0.025, par_e < 0.0132) |>
  select(tmp, d, mh, logg, mag, l) 

# preproc_l <- preProcess(df, method = c("center", "scale"))
# df <- predict(preproc_l, df)
df <- df[1:300,]
coords <- coords[1:300,]

b <- plot_ly(x = coords$x, y = coords$y, z = coords$z, type = "scatter3d", mode = "markers", marker = list(size = 1.5, color = "yellow"))

#3D PLOT OF STARS USED
b <- b %>% layout(
  scene = list(
    bgcolor = 'black',
    xaxis = list(range = c(-1e4, -100), title = "x (pc)"),  # Set x-axis limits
    yaxis = list(range = c(-1100, 6000), title = "y (pc)"),  # Set y-axis limits
    zaxis = list(range = c(-12000, 1300), title = "z (pc)")   # Set z-axis limits
  )
)
b

paste("Range of luminocity values:",min(df$l), max(df$l))
df$l <- (df$l/1e25)
df$logl <- log(df$l)
  
  ggplot(df, aes(x = logl)) + 
    geom_histogram(fill = "blue", color = "black") + 
    labs(title = "Histogram of Log-Luminosity", x = "Values", y = "Frequency")
```

conclusions:
we need to get less than 25000 parsecs to stay within the milkyway
error of ra, dec and parallax now it makes sense (0.05)
there is many many observation lines, as shown as the very straight line
cut interval errors off at 1st quartile: tmp = 24, mh = 0.0476, logg = 0.025
luminocity is an extremely right skewed graph

  

  