```{r}
library(dplyr)
library(scatterplot3d)
library(plotly)
library(car)
library(caret)
set.seed(123)
```

UNITS:
  milliarcseconds
  log(Fe/H)
  Kelvin
  no units

Lets convert the polar coordinates to cartesian
```{r}
data <- read.csv("star_data_coords.csv")
df <- na.omit(data)

colnames(df) <- c("d", "tmp", "mh", "logg", "mag", "ra", "dec", "par")

distance <- 1/(df$par/1000) #divide the parallax by 1000 to convert milliarcseconds to arcseconds

# Convert RA and Dec to radians
ra_rad <- df$ra*pi/180
dec_rad <- df$dec*pi/180

# Calculate x, y, z coordinates
x <- distance * cos(dec_rad) * cos(ra_rad)
y <- distance * cos(dec_rad) * sin(ra_rad)
z <- distance * sin(dec_rad)

coords <- data.frame(x = x, y = y, z = z)

par(bg = "grey10",
    mar = c(5, 5, 2, 2))
scatterplot3d(x, y, z, pch = 16, color = "yellow", 
              xlim = c(-1e6, 1e6), ylim = c(-5e6, 5e6), zlim = c(-4e6, 4e6), 
              cex.symbols = 0.3,
              col.main = "white",   # Title text color
              col.lab = "white",    # Axis labels color
              col.axis = "white",
              main = "Star distance from Earth (pc)")

p <- plot_ly(x = x, y = y, z = z, type = "scatter3d", mode = "markers", marker = list(size = 1.5, color = "yellow"))

# Modify axis limits for 3D plot
p <- p %>% layout(
  scene = list(
    bgcolor = 'black',
    xaxis = list(range = c(-1e6, 1e6)),  # Set x-axis limits
    yaxis = list(range = c(-5e6, 5e6)),  # Set y-axis limits
    zaxis = list(range = c(-4e6, 4e6))   # Set z-axis limits
  )
)
p
```
  

Take errors into account
```{r}
data <- read.csv("star_data_e.csv")
df <- na.omit(data)

colnames(df) <- c("tmp", "mh", "logg", "mag", "ra", "dec", "par", "ra_e", "dec_e", "par_e")



ggplot(df, aes(x = ra_e)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of RA Errors", x = "Error", y = "Frequency") +
  theme_minimal()

ggplot(df, aes(x = dec_e)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Dec Errors", x = "Error", y = "Frequency") +
  theme_minimal()

ggplot(df, aes(x = par_e)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Parallax Errors", x = "Error", y = "Frequency") +
  theme_minimal()



df_no_error <- df |>
  filter(ra_e <0.075, dec_e <0.075, par_e < 0.075)

distance <- 1/(df$par/1000) #divide the parallax by 1000 to convert milliarcseconds to arcseconds

# Convert RA and Dec to radians
ra_rad <- df_no_error$ra*pi/180
dec_rad <- df_no_error$dec*pi/180

# Calculate xyz coordinates
x <- distance * cos(dec_rad) * cos(ra_rad)
y <- distance * cos(dec_rad) * sin(ra_rad)
z <- distance * sin(dec_rad)

coords <- data.frame(x = x, y = y, z = z)

par(bg = "grey10",
    mar = c(5, 5, 2, 2))

scatterplot3d(x, y, z, pch = 16, color = "yellow", 
              xlim = c(-1e6, 1e6), ylim = c(-5e6, 5e6), zlim = c(-4e6, 4e6), 
              cex.symbols = 0.3,
              col.main = "white",   # Title text color
              col.lab = "white",    # Axis labels color
              col.axis = "white",
              main = "Star distance from Earth (pc)")

p <- plot_ly(x = x, y = y, z = z, type = "scatter3d", mode = "markers", marker = list(size = 1.5, color = "yellow"))

# Modify axis limits for 3D plot
p <- p %>% layout(
  scene = list(
    bgcolor = 'black',
    xaxis = list(range = c(-1e6, 1e6)),  # Set x-axis limits
    yaxis = list(range = c(-5e6, 5e6)),  # Set y-axis limits
    zaxis = list(range = c(-4e6, 4e6))   # Set z-axis limits
  )
)
p
```

  
  
  
```{r}
data <- read.csv("star_data_coords.csv")
df <- na.omit(data)

colnames(df) <- c("d", "tmp", "mh", "logg", "mag", "ra", "dec", "par")
df$d <- 1000/df$par

linear <- lm(mh ~ par + tmp + logg + mag, data = df)
summary(linear)
```
***Talk about boring things***

Wow so many significant variables, lets see why
```{r}
#check variance inflation factor (VIF)
vif <- vif(linear)
vif

#normalize data
linear_data <- df |>
  select(mh, par, tmp, logg, mag)

preproc_l <- preProcess(linear_data, method = c("center", "scale"))
linear_data_norm <- predict(preproc_l, linear_data)

linear_norm <- lm(mh ~ par + tmp + logg + mag, data = linear_data_norm)
summary(linear_norm)

#lower sample size
subset_linear <- linear_data_norm |> sample_n(500)
linear_small <- lm(mh ~ par + tmp + logg + mag, data = subset_linear)
summary(linear_small)
```

***Talk about how to pick the right sample size***
  
Try to find how many datapoints are within the milkyway
```{r}
ggplot(df, aes(x = d)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(title = "Density Plot of Distances", x = "Distance (pc)", y = "Density") +
  xlim(0, 100000)

ggplot(df, aes(x = d)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(title = "Density Plot of Distances", x = "Distance (pc)", y = "Density") +
  xlim(0, 5.5e6)
```

Let's remove all data outside of 25,000 parsecs to get only stars within the milkyway
```{r}
data <- read.csv("star_data_e.csv")
df <- na.omit(data)

colnames(df) <- c("tmp", "mh", "logg", "mag", "ra", "dec", "par", "ra_e", "dec_e", "par_e")
df$d <- 1000/df$par

df <- df |> filter(abs(d) < 25000)

df_no_error <- df |>
  filter(ra_e <0.02, dec_e <0.02, par_e < 0.02)

distance <- 1/(df_no_error$par/1000) #divide the parallax by 1000 to convert milliarcseconds to arcseconds

# Convert RA and Dec to radians
ra_rad <- df_no_error$ra*pi/180
dec_rad <- df_no_error$dec*pi/180

# Calculate xyz coordinates
x <- distance * cos(dec_rad) * cos(ra_rad)
y <- distance * cos(dec_rad) * sin(ra_rad)
z <- distance * sin(dec_rad)

b <- plot_ly(x = x, y = y, z = z, type = "scatter3d", mode = "markers", marker = list(size = 1.5, color = "yellow"))

# Modify axis limits for 3D plot
b <- b %>% layout(
  scene = list(
    bgcolor = 'black',
    xaxis = list(range = c(-2e4, 2e4)),  # Set x-axis limits
    yaxis = list(range = c(-2.5e4, 2.5e4)),  # Set y-axis limits
    zaxis = list(range = c(-2e4, 2e4))   # Set z-axis limits
  )
)
b

#Make a pretty graph with earth in it
coords <- data.frame(x, y, z,
  source = rep("dataset", length(x)),
  size = rep(1.5, length(x))# Assign the label 'dataset' to each point
)

df <- df |> filter(abs(d) > 1000)
# Define the extra point (for example, at coordinates 6, 2, 3)
earth <- data.frame(
  x = 0, 
  y = 0, 
  z = 0,
  source = "earth",
  size = 4
  # Label the extra point as 'extra_point'
)

df_with_earth <- rbind(coords, earth)

p <- plot_ly(data = df_with_earth, x = ~x, y = ~y, z = ~z, type = "scatter3d", mode = "markers", marker = list(size = ~size), color = ~source, colors = c("yellow", "deepskyblue2"))

p <- p %>% layout(
  scene = list(
    bgcolor = 'black',
    xaxis = list(range = c(-2e4, 2e4)),  # Set x-axis limits
    yaxis = list(range = c(-2.5e4, 2.5e4)),  # Set y-axis limits
    zaxis = list(range = c(-2e4, 2e4))   # Set z-axis limits
  )
)
p
```


conclusions:
we found sweet spot of 500 stars
we need to get less than 25000 parsecs to stay within the milkyway
error of ra, dec and parallax now it makes sense
  
  
  